It's all about Meaning!\\{\normalsize A Theory for Data Migration during Software Development

### 2022-12-11 2022 Migration/articleMigration.tex
changes made together on dec 11th

\section{Terminology}
\label{sct:Terminology}
   First,
   we define ``information system'' as a combination of dataset, schema, and functionality.
   We start with datasets in section~\ref{sct:Datasets}, then we define schemas in section~\ref{sct:Schemas},
   and then we can define information systems.

@@

   The relation $\inst:\Pair{\Atoms}{\Concepts}$ is the instance relation between atoms and concepts.
   The expression $a\ \inst\ C$ means that atom $a$ is an \emph{instance} of concept $C$.
   This relation is used in the type system, in which $\inst$ assigns a type to every atom.
   This makes $\inst$ a total relation:
\begin{eqnarray}
   \forall a\in\atoms\ \exists C\in\Concepts: a\ \inst\ C
\label{eqn:concepts}
\end{eqnarray}

@@

   Triples serve to represent data.
   A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
\begin{eqnarray}
   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
   \label{eqn:type}
\end{eqnarray}
   For example, $\triple{\text{\atom{Peter}}}{\declare{\id{phone}}{\tt Person}{\tt TelephoneNumber}}{\text{\atom{074238991}}}$ is a triple.
   Equation~\ref{eqn:type} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
   In practice, users can say that Peter has telephone number 074238991.
   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
   This ``meaning from practice'' has no consequences in the formal world.
   We leave it entirely up to a user to attach a practical meaning to a triple.

@@

   A relation $r$ can serve as a container of pairs,
   as defined by the function $\id{pop}_r:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
   It defines a set of pairs, which we call the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples_{\dataset}\}
\label{eqn:pop}
\end{equation}
   Equation~\ref{eqn:type} implies that for every dataset $\dataset$:
\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ \inst\ A\ \wedge\ b\ \inst\ B\]
   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
   This allows for static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.

@@

   A rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
   For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
   that represents the set of violations of rule $u$ in any dataset.
   We also assume a function $\Sign{u} : \Dataset\rightarrow\Pair{\Concepts}{\Concepts}$
   that represents the signature of a rule, i.e. the type of atoms in violations:
\begin{eqnarray}
   \sign{u}{\dataset}=\pair{A}{B}\ \wedge\ \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
   \label{eqn:typed violations}
\end{eqnarray}
   To determine whether rule $u$ is satisfied in dataset $\dataset$,
   we define $\sat{u}{\Dataset}$:
\begin{eqnarray}
   \sat{u}{\dataset}&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
   \label{eqn:sat}
\end{eqnarray}
   Violations are used to produce meaningful error messages and to trigger changes in the dataset to restore satisfaction of that rule.

@@

   %These rules reflect the meaning (semantics) that is shared among users.

\subsection{Functionality}
   The functionality of an information system is an integral part of it.
   It consists of software.
   In this paper, we are not interested in the precise nature of the functionality.

\subsection{Information Systems}
\label{sct:Information Systems}
   This section defines the notion of information system.
   As before, we will suffix the elements of this definition if the situation requires it.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\functionality,\roles,\maintain\ra$, in which
\begin{itemize}
   \item dataset $\dataset$ is defined as in section~\ref{sct:Datasets};
   \item schema $\schema$ is defined as in section~\ref{sct:Schemas};
   \item $\functionality$ represents the functionality of the system, which is typically software;
   \item $\roles$ is a set of roles;
   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$.
\end{itemize}
\end{definition}
   A \define{role} is a name that identifies a group of users.
   It serves as a placeholder for a person or a machine (i.e. an actor) who can change the dataset.
   The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
   In the sequel, when we talk about a role we actually mean an actor who fulfills that role in the system.

@@

   The example system also has a second relation that states which modules are part of which course.
   It is constrained by \verb-[UNI]- to be univalent.
   This means that every \verb-Module- is part of at most one \verb-Course-.
\begin{verbatim}
RELATION isPartOf[Module*Course] [UNI] =
[ ("Finance", "Management")
; ("Business Rules", "Business IT")
; ("Business Analytics", "Business IT")
; ("IT-Governance", "Management")
]
\end{verbatim}
   The third and last relation states which students are enrolled for which module.
   We leave it empty for now.
\begin{verbatim}
RELATION isEnrolledFor[Student*Module]
\end{verbatim}
   We also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that a student takes.
   In Ampersand, which is a syntactically sugared form of relation algebra,
   we give each rule a name and declare a role to preserve its invariance:
\begin{verbatim}
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
ROLE Administrator MAINTAINS EnrollRule
\end{verbatim}
   The actual constraint, \verb#isEnrolledFor |- takes;isPartOf~#, is written in logic as:
\begin{equation}
   \begin{array}{l}\forall s\in\text{\tt Student}, m\in\text{\tt Module}\ \exists c\in\text{\tt Course}:\\
s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
\end{array}
\label{eqn:example isEnrolledFor}
\end{equation}

   This example contains a dataset with seven triples
   and a schema with three concepts, three relations, and one rule.
   Now let us check the requirements to verify that this example defines an information system.
   The Ampersand compiler generates a dataset $\dataset$, which contains a set of triples and a relation $\inst$.
   It defines the set of triples $\triples$ as:
   \[\begin{array}[t]{l}
         \triple{\text{\tt "Peter"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
         \triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "Finance"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
         \triple{\text{\tt "Business Rules"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "Business Analytics"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "IT-Governance"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Management"}}
   \end{array}\]
   The relation $\id{inst}$ contains the pairs:
   \[\begin{array}{l}
      \pair{\tt "Finance"}{\tt Module}\\
      \pair{\tt "Business Rules"}{\tt Module}\\
      \pair{\tt "Business Analytics"}{\tt Module}\\
      \pair{\tt "IT-Governance"}{\tt Module}\\
      \pair{\tt "Management"}{\tt Course}\\
      \pair{\tt "Business IT"}{\tt Course}\\
      \pair{\tt "Peter"}{\tt Student}\\
      \pair{\tt "Susan"}{\tt Student}\\
      \pair{\tt "John"}{\tt Student}
   \end{array}\]
   Requirement~\ref{eqn:concepts} is satisfied because every atom has a concept.
   Dataset $\pair{\triples}{\inst}$ satisfies both requirements~\ref{eqn:type} and~\ref{eqn:pop}.
   So, the dataset $\dataset=\pair{\triples}{\inst}$ satisfies the requirements from section~\ref{sct:Datasets}

@@

   To fulfill the definition of information system, definition~\ref{def:information system},
   Ampersand generates a set of roles $\roles=\{{\tt Administrator}\}$ and
   a relation $\maintain$, which contains one pair only: $\pair{{\tt Administrator}}{{\tt EnrollRule}}$.
   For the purpose of this paper, the functionality $\functionality$ is irrelevant (onderbouwen!).
   The constraint~\ref{eqn:example isEnrolledFor} of rule {\tt EnrollRule} is satisfied
   because relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty.
   So, there are no violations, which satisfies requirement~\ref{eqn:typed violations}.
   With only one rule in the system, requirement~\ref{eqn:sat} is satisfied too.
   Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
   This completes the requirements for the entire quintuple $\la\dataset,\schema,\functionality,\roles,\maintain\ra$.
   So we can conclude that information system $\infsys$ satisfies definition~\ref{def:information system}.

%    The language Ampersand is typed, to prevent a substantial amount of programming mistakes at compile-time.
%    This means that every atom is an instance of a concept:
% \begin{eqnarray}
%    \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:concepts}
% \end{eqnarray}
%    To do type checking based on the schema only,
%    all relations in triples from the dataset must be ``known'' in the schema:
% \begin{equation}
%    \forall a,r,b:\ \triple{a}{r}{b}\in\triples\ \Rightarrow\ r\in\rels\label{eqn:define R}
% \end{equation}

%    In a typed information system, atoms in a relation have a type as declared in a relation.
%    For this reason, every relation $r$ has not only a name, but also a source concept and a target concept.
%    We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
% \begin{eqnarray}
%    \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\label{eqn:concepts}\\
%    \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B\label{eqn:type}
% \end{eqnarray}
%    The type checker in the Ampersand compiler will spot violations and refuse to generate a system if they occur.

>>>>

\section{Terminology}
\label{sct:Terminology}
   An ``information system'' is a combination of dataset, schema, and functionality.
   For the purpose of this paper, we ignore the functionality because it does not impact the migration.
   We start with datasets in section~\ref{sct:Datasets}, then we define schemas in section~\ref{sct:Schemas},
   and then we can define information systems.

@@

   The relation $\inst:\Pair{\Atoms}{\Concepts}$ is the instance relation between atoms and concepts.
   The expression $a\ \inst\ C$ means that atom $a$ is an \emph{instance} of concept $C$.
   This relation is used in the type system, in which $\inst$ assigns a type to every atom.
   This makes $\inst$ total on the atoms in the dataset:
\begin{eqnarray}
   \forall a\in\atoms\ \exists C\in\Concepts: a\ \inst\ C
\label{eqn:typedAtom}
\end{eqnarray}

@@

   Triples serve to represent data.
   A triple is an element of $\Triple{\Atoms}{\Rels}{\Atoms}$.
   A dataset $\dataset$ is a tuple $\pair{\triples}{\inst}$ that satisfies:
\begin{eqnarray}
   \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B
   \label{eqn:wellTypedEdge}
\end{eqnarray}
   For example, $\triple{\text{\atom{Peter}}}{\declare{\id{phone}}{\tt Person}{\tt TelephoneNumber}}{\text{\atom{074238991}}}$ is a triple.
   Equation~\ref{eqn:wellTypedEdge} says that \atom{Peter} is an instance of {\tt Person} and \atom{074238991} is an instance of {\tt TelephoneNumber}.
   In practice, users can say that Peter has telephone number 074238991.
   So, the ``thing'' that \atom{Peter} refers to (which is Peter) has \atom{074238991} as a telephone number.
   This ``meaning from practice'' has no consequences in the formal world.
   We leave it entirely up to a user to attach a practical meaning to a triple.

@@

   A relation $r$ can serve as a container of pairs,
   as defined by the function $\id{pop}_r:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$.
   It defines a set of pairs, which we call the population of $r$:
\begin{equation}
   \pop{r}{\dataset}\ =\ \{ \pair{a}{b}\mid\ \triple{a}{r}{b}\in\triples_{\dataset}\}
\label{eqn:pop}
\end{equation}
   Equation~\ref{eqn:wellTypedEdge} implies that for every dataset $\dataset$:
\[\pair{a}{b}\in\pop{\declare{n}{A}{B}}{\dataset}\ \Rightarrow\ \inst\ A\ \wedge\ b\ \inst\ B\]
   For a developer, this means that the type of an atom depends only on the relation in which it resides; not on the actual population of the database.
   This allows for static typing, which has well established advantages for the software engineering process~\cite{HanenbergKRTS14,Petersen2014}.

@@

   A rule in a schema serves to constrain the dataset, to ensure its semantic integrity.
   For every rule $u$ in the schema, we assume there is a function $\id{viol}_u : \Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$
   that represents the set of violations of rule $u$ in any dataset.
   We also assume a function $\Sign : \rules\rightarrow\Pair{\Concepts}{\Concepts}$
   that represents the signature of a rule, i.e. the type of atoms in violations:
\begin{eqnarray}
   \sign{u}=\pair{A}{B}\\
   \pair{a}{b}\in\viol{u}{\dataset}&\Rightarrow&a\ \inst\ A\wedge b\ \inst\ B
   \label{eqn:wellTypedEdged violations}
\end{eqnarray}
   To determine whether rule $u$ is satisfied in dataset $\dataset$,
   we define $\sat{u}{\Dataset}$:
\begin{eqnarray}
   \sat{u}{\dataset}&\Leftrightarrow&\viol{u}{\dataset}=\emptyset
   \label{eqn:sat}
\end{eqnarray}
   Violations are used to produce meaningful error messages and to trigger changes in the dataset to restore satisfaction of that rule.

@@

\subsection{Information Systems}
\label{sct:Information Systems}
   This section defines the notion of information system.
   As before, we will suffix the elements of this definition if the situation requires it.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\la\dataset,\schema,\roles,\maintain\ra$, in which
\begin{itemize}
   \item dataset $\dataset$ is defined as in section~\ref{sct:Datasets};
   \item schema $\schema$ is defined as in section~\ref{sct:Schemas};
   \item $\roles$ is a set of roles;
   \item $\maintain : \roles\times\rules$ is the maintainance relation between roles and rules from $\schema$.
\end{itemize}
\end{definition}
   A \define{role} is a name that identifies a group of users.
   It serves as a placeholder for a person or a machine (i.e. an actor) who can change the dataset.
   The purpose of a role is to mention an individual user (human) or an automated actor (bot) without knowing who that user is.
   In the sequel, when we talk about a role we actually mean an actor who fulfills that role in the system.

@@

   The example system also has a second relation that states which modules are part of which course.
   It is constrained by \verb-[UNI]- to be univalent.
   This means that every \verb-Module- is part of at most one \verb-Course-.
\begin{verbatim}
RELATION isPartOf[Module*Course] [UNI] =
[ ("Finance", "Management")
; ("Business Rules", "Business IT")
; ("Business Analytics", "Business IT")
; ("IT-Governance", "Management")
]
\end{verbatim}
   The third and last relation states which students are enrolled for which module.
   We leave it empty for now.
\begin{verbatim}
RELATION isEnrolledFor[Student*Module]
\end{verbatim}
   We also define a rule, {\tt EnrollRule}, that states that a student can enroll for any module that is part of a course that a student takes.
   In Ampersand, which is a syntactically sugared form of relation algebra,
   we give each rule a name and declare a role to preserve its invariance:
\begin{verbatim}
RULE EnrollRule: isEnrolledFor |- takes;isPartOf~
ROLE Administrator MAINTAINS EnrollRule
\end{verbatim}
   The semantics of Ampersand defines $\viol{\tt EnrollRule}{\dataset}$ as follows:
\begin{equation}
   \viol{\tt EnrollRule}{\dataset}\ =\{\begin{array}{l}
   \pair{m}{s}\mid\neg\exists c\in\text{\tt Course}:\\
s\ \text{\tt isEnrolledFor}\ m\ \rightarrow\ s\ \text{\tt takes}\ c\ \wedge\ m\ \text{\tt isPartOf}\ c
\end{array}\}
\label{eqn:example isEnrolledFor}
\end{equation}

   This example contains a dataset with seven triples
   and a schema with three concepts, three relations, and one rule.
   Now let us check the requirements to verify that this example defines an information system.
   The Ampersand compiler generates a dataset $\dataset$, which contains a set of triples and a relation $\inst$.
   It defines the set of triples $\triples$ as:
   \[\begin{array}[t]{l}
         \triple{\text{\tt "Peter"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
         \triple{\text{\tt "Susan"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "John"}}{\declare{\text{\tt takes}}{\text{\tt Student}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "Finance"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Management"}}\\
         \triple{\text{\tt "Business Rules"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "Business Analytics"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Business IT"}}\\
         \triple{\text{\tt "IT-Governance"}}{\declare{\text{\tt isPartOf}}{\text{\tt Module}}{\text{\tt Course}}}{\text{\tt "Management"}}
   \end{array}\]
   The relation $\id{inst}$ contains the pairs:
   \[\begin{array}{l}
      \pair{\tt "Finance"}{\tt Module}\\
      \pair{\tt "Business Rules"}{\tt Module}\\
      \pair{\tt "Business Analytics"}{\tt Module}\\
      \pair{\tt "IT-Governance"}{\tt Module}\\
      \pair{\tt "Management"}{\tt Course}\\
      \pair{\tt "Business IT"}{\tt Course}\\
      \pair{\tt "Peter"}{\tt Student}\\
      \pair{\tt "Susan"}{\tt Student}\\
      \pair{\tt "John"}{\tt Student}
   \end{array}\]
   Requirement~\ref{eqn:typedAtom} is satisfied because every atom has a concept.
   The tuple $\pair{\triples}{\inst}$ satisfies both requirements~\ref{eqn:wellTypedEdge} and~\ref{eqn:pop}.
   So, this defines a dataset $\dataset=\pair{\triples}{\inst}$ as introduced in section~\ref{sct:Datasets}.

@@

   To fulfill the definition of information system, definition~\ref{def:information system},
   Ampersand generates a set of roles $\roles=\{{\tt Administrator}\}$ and
   a relation $\maintain$, which contains one pair only: $\pair{{\tt Administrator}}{{\tt EnrollRule}}$.
   The constraint~\ref{eqn:example isEnrolledFor} of rule {\tt EnrollRule} is satisfied
   because relation $\declare{\tt isEnrolledFor}{\tt Student}{\tt Module}$ is empty.
   So, there are no violations, which satisfies requirement~\ref{eqn:wellTypedEdged violations}.
   With only one rule in the system, requirement~\ref{eqn:sat} is satisfied too.
   Requirement~\ref{eqn:specialization} is satisfied because this example contains no specialization.
   This completes the requirements for $\la\dataset,\schema,\roles,\maintain\ra$.
   So we can conclude that information system $\infsys$ satisfies definition~\ref{def:information system}.

%    The language Ampersand is typed, to prevent a substantial amount of programming mistakes at compile-time.
%    This means that every atom is an instance of a concept:
% \begin{eqnarray}
%    \forall a\in\atoms\ \exists A\in\concepts&:&a\ \inst\ A\label{eqn:typedAtom}
% \end{eqnarray}
%    To do type checking based on the schema only,
%    all relations in triples from the dataset must be ``known'' in the schema:
% \begin{equation}
%    \forall a,r,b:\ \triple{a}{r}{b}\in\triples\ \Rightarrow\ r\in\rels\label{eqn:define R}
% \end{equation}

%    In a typed information system, atoms in a relation have a type as declared in a relation.
%    For this reason, every relation $r$ has not only a name, but also a source concept and a target concept.
%    We write $r=\declare{nm}{A}{B}$ to denote that relation $r$ has name \id{nm}, source concept $A$, and target concept $B$.
% \begin{eqnarray}
%    \forall\declare{n}{A}{B}\in\rels&:&A\in\concepts\ \wedge\ B\in\concepts\label{eqn:typedAtom}\\
%    \forall\triple{a}{\declare{n}{A}{B}}{b}\in\triples&:&a\ \inst\ A\ \wedge\ b\ \inst\ B\label{eqn:wellTypedEdge}
% \end{eqnarray}
%    The type checker in the Ampersand compiler will spot violations and refuse to generate a system if they occur.
